A Data-Driven Framework for Neural Field Modeling

### 2011-01-05 ltx/first_order_framework/response_to_reviews.tex

    \begin{enumerate}
        \item The discrepancy of the results in Experiment II where the weights of the basis functions are not in agreement with the actual parameters while the reconstructed kernels seem satisfying.

        \item The use of Euler discretisation and how this affects the system's behavior.

		\emph{This point was made by a reviewer from our PLoS attempt. The reviewer stated: ``1. The authors might want to reference prior work in using Kalman
		Filtering for the estimation of neural mass models from data [1-3],
		[13]. In fact from those papers stressed the importance of carrying out
		a high order discretization of the differential equations, which should
		be in fact treated as stochastic. In equation (S1.1) of the extended
		derivation, it is shown that an Euler approximation is taken which may
		even change the dynamical properties of the system. The authors should
		discuss this issue.'' ALL THESE PAPERS USE a method called local linearisation approach for discretisation.}

		\emph{We should add some more to the modeling assumptions section to address this one. I have begun to talk about the LL method.}

@@

        \item The derivation of Equations (30) and (31) [spacing of basis functions] and make them more transparent for the general audience.

\emph{To address this point, we have added an extra appendix with a derivation of the sampling theorem with the appropriate cross-referencing to the main text. An effort was made to keep this derivation minimal and as clear as possible. We have also added the references to relevant part of the text (that should have been there before, as the reviewer has correctly pointed out). We felt that it was worth adding the extra appendix to provide a clear derivation that is contextually and notationally relevant to this paper, as the derivations cited papers may not be clear to the Neuroimage general audience.}

        \item The existing literature on inversion schemes for similar models of brain activity.

\emph{The existing literature has been discussed in more depth in the introduction.}

        \item In section 3, Table 6 is mentioned however there is no such table in the ms.

\emph{The cross-referencing has been corrected.}

### 2011-01-05 ltx/first_order_framework/LaTeX_Manuscript.tex
editted cor derivation
add background on estimation schemes suggested by reviewer 1

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the \citet{Wilson1973} (WC) or \citet{Amari1977} models, or a neural mass model like the \citet{Jansen1995} model, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this task has been approached from a number of standpoints \citep{Nunez2000,Jirsa2002,Robinson2004,Daunizeau2009}.

Perhaps the most accepted estimation framework for neural mass models is dynamical causal modeling (DCM) \citep{David2003,David2006}, which has been proposed for studying evoked potential dynamics. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the model of \citet{Jansen1995}. Another recent approach estimates the parameters of a neural field model using an unscented Kalman filter based on modified WC neural field equations \citep{schiff2008kalman}. This work takes a systems theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable.

@@

	R_{y_{t},y_{t+1}}(\boldsymbol{\tau}) &\approx \mathbf{E}\left[\mathcal{F}^{-1} \left\{M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \ast \left(\xi e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} V(\boldsymbol{\nu}) + T_s e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} W(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} V(\boldsymbol{\nu}) \right) \right\} \right] \\
	&= \mathbf{E}\left[\mathcal{F}^{-1} \left\{M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) \ast \left(\xi e^{2i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) + T_s e^{2i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} W(\boldsymbol{\nu}) \right) \right\} \right] \\
	&= \mathbf{E}\left[\mathcal{F}^{-1} \left\{M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) \ast e^{2i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) \left(\xi + T_s e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} W(\boldsymbol{\nu}) \right) \right\} \right] \\

@@

\dean{An advantage in using a mean field model over a more detailed mathematical description is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into the connectivity kernel coefficients that effectively describe the net system gain. In this way a minimal number of assumptions are made. A major challenge in data assimilation for the mass action of the brain is to make models sufficiently detailed such that the parameters are meaningful, but simple enough to yield clear insights that can be related to theoretical studies (from both neuroscience and engineering perspectives).}

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see \citet{Ullah2010} for a nice example). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. A real-time implementation can also be achieved using the UKF (without the smoother) and a real-time LS estimator (see REF for example). }

>>>>

For models to be clinically viable, they must be patient-specific. A possible approach to achieve this would be to fit a general continuum neural field model, like the \citet{Wilson1973} (WC) or \citet{Amari1977} models, or a neural mass model like the \citet{Jansen1995} model, to patient-specific data. Fitting the neural models to individuals is a highly non-trivial task. Recently, however, this task has been approached from a number of standpoints. \dean{(Reviewer 1, additional comment d):) The first paper on patient specific modeling came from~\citet{Valdes1999}, where they fit a neural mass model of \citet{LopesDaSilva1976} and \citet{Zetterberg1978} using an extended Kalman filter. This paper demonstrated for the first time the feasibility of assimilating data with neural models of EEG.}

Perhaps the most accepted estimation framework for neural mass models is dynamical causal modeling (DCM) \citep{David2003,David2006}, which has been proposed for studying evoked potential dynamics. This framework can be viewed as an extension to the work of~\citet{Valdes1999} allowing for coupling of neural masses. Via a Bayesian inference scheme, DCM estimates the long range (cortico-cortical) connectivity structure between the specific isolated brain regions that best explains a given data set using the model of \citet{Jansen1995}.

\dean{(Reviewer 1, additional comment d):) Data-driven neural modeling was extended to the continuum approximation by \citet{Galka2008}, where they proposed an estimation framework based on a linear damped wave equation. Using a Kalman filter maximum likelihood framework, they improved on the low resolution electromagnetic tomography (LORETA) method for solving the inverse EEG problem. Another continuum neural field model-based estimator was developed by \citet{Daunizeau2009}, as an extension to the DCM framework. They argued that this approach is superior for data assimilation than the neural mass DCM, since field models can explain a richer repertoire of dynamics such as traveling waves and bump solutions.}

\dean{Another recent approach estimates the parameters of a modified WC neural field model using an unscented Kalman filter~\citep{schiff2008kalman}. This work takes a systems theoretic approach to the neural estimation problem, successfully demonstrating that it is possible to perform state estimation and control on spatiotemporal neural fields. This marks the first step in what has the potential to revolutionize the treatment of many neurological diseases where therapeutic electrical stimulation is viable. For other valuable contributions to data-driven neural modeling see \citet{Nunez2000,Jirsa2002,Robinson2004}.}

@@

	R_{y_{t},y_{t+1}}(\boldsymbol{\tau}) &\approx \mathbf{E}\left[\mathcal{F}^{-1} \left\{M(\boldsymbol{\nu})V(\boldsymbol{\nu}) \ast \left(\xi e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} V(\boldsymbol{\nu}) + T_s e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} W(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} \varsigma V(\boldsymbol{\nu}) \right) \right\} \right] \\
	&= \mathbf{E}\left[\mathcal{F}^{-1} \left\{M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) \ast \left(\xi e^{2i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) + T_s e^{2i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) \varsigma V(\boldsymbol{\nu}) e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} W(\boldsymbol{\nu}) \right) \right\} \right] \\
	&= \mathbf{E}\left[\mathcal{F}^{-1} \left\{M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) \ast e^{2i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} M(\boldsymbol{\nu}) V(\boldsymbol{\nu}) \left(\xi + T_s \varsigma e^{i\boldsymbol{\nu}^{\top}\boldsymbol{\tau}} W(\boldsymbol{\nu}) \right) \right\} \right] \\

@@

\dean{An advantage in using a lumped-parameter field model over a more detailed mathematical description is that the myriad of parameters that might influence excitability, such as specific ion concentrations, are considered to be lumped into the connectivity kernel coefficients that effectively describe the net system gain. In this way a minimal number of assumptions are made. A major challenge in data assimilation for the mass action of the brain is to make models sufficiently detailed such that the parameters are meaningful, but simple enough to yield clear insights that can be related to theoretical studies (from both neuroscience and engineering perspectives).}

\dean{The framework assumes that the parameters of the model (and brain) are stationary over the estimation period. This is a reasonable assumption if applying the framework to an evoked-potential experiment over a short time period in a controlled environment. However, it is well known that factors that influence excitability are non-stationary over longer time periods (see \citet{Ullah2010} for a nice example). Therefore, the estimation framework must be altered if it is to be applied to data over extended periods of time. A possible way to achieve this is to use a more sophisticated model that accounts for longer term time dependancies (higher temporal order). Alternatively, the parameters of the model could be considered time-varying, where changes in excitability might result in different connectivity kernel parameters. A possible way to achieve this would be to employ window-based estimation, where the estimation is performed on successive windows of data (possibly overlapping windows). The window length will need to be sufficiently short such that the parameters are approximately stationary of the estimation duration, and sufficiently long to achieve a good estimation result. If this trade-off can not be successfully achieved then the level of detail in the model must be increased. A real-time implementation can also be achieved using the UKF (without the smoother) and a real-time LS estimator (see REF for example). }

\dean{We have assumed that the model can be discretized by the first order Euler method. This discretization may change the properties of the stochastic nonlinear system. An alternative is to use the LL method....}
